Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_histogram()
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}

```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}

```

***

### Cut a Variable
Notes:
# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.
# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]
```{r Cut a Variable}
pf <- read.delim('../pseudo_facebook.tsv')
pf$year_joined <- floor(2014 - pf$tenure / 365)
pf$year_joined.bucket <- cut(pf$year_joined, c(2004, 2009, 2011, 2012, 2014))
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```

***

### Plot the Grand Mean
Notes:
# (1) Add another geom_line to code below
# to plot the grand mean of the friend count vs age.

# (2) Exclude any users whose year_joined.bucket is NA.

# (3) Use a different line type for the grand mean.

# As a reminder, the parameter linetype can take the values 0-6:

# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

```{r Plot the Grand Mean}
ggplot(data = subset(pf, !is.na(year_joined.bucket)), 
       aes(x=age, y=friend_count)) +
  geom_line(aes(color = year_joined.bucket),
            stat='summary',
            fun.y = median)
```

***

### Friending Rate
Notes:
What is the median rate?
What is the max rate?
```{r Friending Rate}
pf.tenure <- pf[pf$tenure >= 1, ]
pf.tenure$friend_rate <- pf.tenure$friend_count / pf.tenure$tenure
summary(pf.tenure$friend_rate)
# alternative
with(subset(pf, tenure >= 1), summary(friend_count / tenure))
```

***

### Friendships Initiated
Notes:
# Create a line graph of mean of friendships_initiated per day (of tenure)
# vs. tenure colored by year_joined.bucket.
# You need to make use of the variables tenure,
# friendships_initiated, and year_joined.bucket.
# You also need to subset the data to only consider user with at least
# one day of tenure.

What is the median friend rate?
What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(data = subset(pf, tenure >= 1), 
       aes(x = tenure, y = friendships_initiated / tenure)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_smooth(aes(color = year_joined.bucket))

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv("./yogurt.csv", header = TRUE)
yo$id <- factor(yo$id)
qplot(data = yo, x = price)
hist(yo$price, breaks = seq(0, 100, by = 1))
```

***

### Number of Purchases
Notes:
# Create a new variable called all.purchases,
# which gives the total counts of yogurt for
# each observation or household.

# One way to do this is using the transform
# function. You can look up the function transform
# and run the examples of code at the bottom of the
# documentation to figure out what it does.

```{r Number of Purchases}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + 
            plain + mixed.berry)
```

***

### Prices over Time
Notes:
# Create a scatterplot of price vs time.
# This will be an example of a time series plot.
```{r Prices over Time}
ggplot(data = yo, mapping = aes(x = time, y = price)) +
  geom_point()
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(1234)
sample.ids <- sample(levels(yo$id), 16)
ggplot(data = subset(yo, id %in% sample.ids), mapping = aes(x = time, y = price)) +
  facet_wrap( ~ id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

***
```{r}
install.packages("GGally")
library(GGally)
theme_set(theme_minimal(20))
set.seed(1234)
pf_subset <- pf[, c(2:15)]
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])
```
### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

