Lesson 3
========================================================

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
library(ggplot2)
setwd("F:/udacity/4 ExploreAndSummarizeData/lesson3/")
list.files()
pf <- read.csv("pseudo_facebook.tsv", sep = "\t")
str(pf)
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
library(ggplot2)
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)
```

***

#### What are some things that you notice about this histogram?

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}

```

#### Let's take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x = friend_count, data = pf)
```

#### How is this plot similar to Moira's first plot?
Response:
The plot is showing the same pattern as Moira's first plot, having the majority of 
users having a low number of friend count (less than 1000) and serveral extreme cases above 4000 friends. Because the largest number of friends is less than 5000, thus it 
is good data in an extreme case.
***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x = friend_count, data = pf, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf,
      color = I('white'), fill = I('#099DD9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, binwidth = 1,
      color = I('white'), fill = I('#099DD9')) + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf, binwidth = 1,
      xlab = "Tenure by year", 
      ylab = "Number of user",
      color = I('white'), fill = I('#099DD9')) + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
```

***

### User Ages
Notes:

```{r User Ages}
qplot(x = age, data = pf, binwidth = 1,
      color = I('white'), fill = I('#099DD9')) +
  scale_x_continuous(breaks = seq(0, 110, 10))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
```{r}
qplot(x = friend_count, data = pf)
summary(pf$friend_count)
summary(log10(pf$friend_count+1))
summary(sqrt(pf$friend_count))
```


Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
library(gridExtra)
# g0 = qplot(x = friend_count, data = pf)
g0 = ggplot(aes(x = friend_count), data = pf) + geom_histogram()
g1 = g0 + scale_x_log10()
g2 = g0 + scale_x_sqrt()
grid.arrange(g0, g1, g2, ncol = 2)
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))

ggplot(aes(x = friend_count, y = ..count../sum(..count..)), 
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend count') +
  ylab('Percentage of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
ggplot(aes(x=www_likes), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) +
  scale_x_log10()
```
### Likes on the Web solution
```{r}
by(pf$www_likes, pf$gender, sum)
```

***

### Box Plots
Notes:

```{r Box Plots}
qplot(x=gender, y=friend_count, 
      data=subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  scale_y_continuous(limits = c(0, 1000))

qplot(x=gender, y=friend_count, 
      data=subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))

by(pf$friend_count, pf$gender, summary)
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, Who made more friend requests - men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
Calculate the total number of users who initiated friendships.
This is reflected in the data by counting the users whose friendships_initiated > 0.
```{r Friend Requests by Gender}
qplot(x = gender, y = friendships_initiated, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 200))

by(pf$friendships_initiated, pf$gender, summary)

```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)
pf$mobile_check_in <- NA

pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

```

What percent of check in using mobile?

Response:
```{r}
sum(pf$mobile_check_in == 1)/length(pf$mobile_check_in)
```
***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!
