Lesson 4
========================================================

***
```{r}
library(ggplot2)
load("./.RData")
```


### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
qplot(x = age, y = friend_count, data = pf)
```

***

#### What are some things that you notice right away?
Response:
1. some outliers in age (>90)
2. most of the users' friend count is below 1000
3. users whose age under 30, the friend count varies more than users above 30
***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 0.05) +
  xlim(13, 90)
```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 0.05) +
  xlim(13, 90)
```

#### What do you notice in the plot?
Response:
Friend count reaches a peak around 20 and start decreasing. There is also a peak 
at around 70.
***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 0.05) +
  xlim(13, 90)
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 0.05, position = position_jitter(height = 0)) +
  xlim(13, 90) +
  coord_trans(y = 'sqrt')
```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(data = pf, aes(x = age, y = friendships_initiated)) +
  geom_point(alpha = 0.05, position = position_jitter(height = 0)) +
  coord_trans(y = 'sqrt')
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)
age_group <- group_by(pf, age)
pf.fc_by_age <- summarize(age_group,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n())
```
```{r another solution}
pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)

head(pf.fc_by_age)
```

Create your plot!
```{r Conditional Means Plot}
# Plot mean friend count vs. age using a line graph.
# Be sure you use the correct variable names
# and the correct data frame. You should be working
# with the new data frame created from the dplyr
# functions. The data frame is called 'pf.fc_by_age'.

# Use geom_line() rather than geom_point to create
# the plot. You can look up the documentation for
# geom_line() to see what it does.
ggplot(data = pf.fc_by_age, aes(x = age, y = friend_count_mean)) +
  geom_line()

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  coord_cartesian(xlim = c(13, 90)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
#  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9),
            linetype = 2, color = 'blue')
```


#### What are some of your observations of the plot?
Response:
Having more than 1000 friends is quite rare. The 90% tile is below 1000.

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(x = pf$age, y = pf$friend_count,
         method = 'pearson')
# alternatives
with(pf, cor.test(age, friend_count, method = 'pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count, method = 'pearson'))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
# Create a scatterplot of likes_received (y)
# vs. www_likes_received (x). Use any of the
# techniques that you've learned so far to
# modify the plot.

ggplot(data = pf, aes(x = www_likes_received+1, y = likes_received+1)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(data = pf, aes(x = www_likes_received, y = likes_received)) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(www_likes_received, likes_received))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
data(Mitchell)
```

Create your plot!

```{r Temp vs Month}
ggplot(data = Mitchell, aes(x = Month, y = Temp)) +
  geom_line()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.
0
b. What is the actual correlation of the two variables?
(Round to the thousandths place)
0.057
```{r Noisy Scatterplots}
with(Mitchell, cor.test(Month, Temp))
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(data = Mitchell, aes(x = Month, y = Temp)) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, max(Mitchell$Month), 12))
```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(data = pf.fc_by_age,
       aes(x = age, y = friend_count_mean)) +
  geom_line()

head(pf.fc_by_age, 10)
pf.fc_by_age[17:19, ]
```

***

### Age with Months Means

```{r Age with Months Means}
# Create a new variable, 'age_with_months', in the 'pf' data frame.
# Be sure to save the variable in the data frame rather than creating
# a separate, stand-alone variable. You will need to use the variables
# 'age' and 'dob_month' to create the variable 'age_with_months'.

# Assume the reference date for calculating age is December 31, 2013.
month_count <- function(date){
  lt <- as.POSIXlt(as.Date(date, origin="1900-01-01"))
  lt$year*12 + lt$mon
}

month_diff <- function(date1, date2) {
  month_count(date1) - month_count(date2)
}

pf$dob <- with(pf, paste(dob_year, dob_month, dob_day, sep = '-'))
pf$age_with_months <- month_diff('2013-12-31', pf$dob) / 12
  

```

Programming Assignment
```{r Programming Assignment}
# Create a new data frame called
# pf.fc_by_age_months that contains
# the mean friend count, the median friend
# count, and the number of users in each
# group of age_with_months. The rows of the
# data framed should be arranged in increasing
# order by the age_with_months variable.

# For example, the first two rows of the resulting
# data frame would look something like...

# age_with_months  friend_count_mean	friend_count_median	n
#              13           275.0000                  275 2
#        13.25000           133.2000                  101 11
library(dplyr)
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarize(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
# Create a new scatterplot showing friend_count_mean
# versus the new variable, age_with_months. Be sure to use
# the correct data frame (the one you create in the last
# exercise) AND subset the data to investigate
# users with ages less than 71.

ggplot(data = subset(pf.fc_by_age_months, age_with_months < 71),
       aes(x = age_with_months, 
           y = friend_count_mean)) +
  geom_line()
  
```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

library(gridExtra)
p1 <- ggplot(data = subset(pf.fc_by_age, age < 71),
             aes(x = age, 
                 y = friend_count_mean)) +
  geom_line() +
  geom_smooth()

p2 <- ggplot(data = subset(pf.fc_by_age_months, age_with_months < 71),
             aes(x = age_with_months, 
                 y = friend_count_mean)) +
  geom_line() +
  geom_smooth()

p3 <- ggplot(data = subset(pf, age < 71),
             aes(x = round(age / 5)*5, y = friend_count)) +
  geom_line(stat = 'summary', fun.y = mean)

grid.arrange(p3, p2, p1, ncol = 1)
```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

