Loan analysis by Xin Pang
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

packages <- c("dplyr", "ggplot2", "gridExtra", "plyr")
new.packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new.packages)){
  install.packages(new.packages)
}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(plyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df.loan <- read.csv("./prosperLoanData.csv", header = TRUE)
```

**Background**
Prosper.com is a peer-to-peer lending website privoding individuals with 
services to borrow money or invest in personal loans. According to [Wikipedia](https://en.wikipedia.org/wiki/Prosper_Marketplace) this 
company used two different ways to calculate loan interest rates. From 2006 to 
2009 the loan interest rate was determined by borrowers and lenders in a bidding
system. Afterwards the rate is predefined by Prosper based on a formula 
evaluating each prospective borrower's credit risk. The dataset to be analyzed
contains users' loan details from November 2005 to March 2014, including 113937 
loan records with 81 variables.

# Overview

We want to know the type of each column, wether they are numeric or factor, 
such that we can select a few of them for analysis. Due to the large number of 
variables in the dataframe, here I only show the type of first 10 attributes.
From the result we can see there are both factor variables and numerical variables.

```{r echo=FALSE, Univariate_Plots_0}
head(lapply(df.loan, class), n = 10)
```

The main features of interest are the following:

* Term: The length of the loan expressed in months. (integer->factor)
* Loan status: The current status of the loan. (factor)
* Borrower rate: The borrower's interest rate for this loan. (numeric)
* Listing category: The category of the listing that the borrower selected when 
posting their listing. (integer->factor)
* Credit score range lower: The lower value representing the range of the borrower's credit score as provided by a consumer credit rating agency. (numeric)
* Credit score range upper: The upper value representing the range of the borrower's credit score as provided by a consumer credit rating agency. (numeric)
* Income range: The income range of the borrower at the time the listing was 
created. (factor)
* Loan original amount: The origination amount of the loan. (integer)
* Loan origination date: The origination date of the loan. (factor->date)
* Percent funded: Percent the listing was funded. (numeric)
* Recommendations: Number of recommendations the borrower had at the time the 
listing was created. (integer)
* Investors: The number of investors that funded the loan. (integer)

There are a few variables requiring a conversion for proper types.

* Term: contains 3 values and can be changed to a factor.
* Listing category: they are actually coded into numbers, each number 
representing one category, thus it should be factor instead of integer.
* Loan origination date: should be in date format.

Below processes the column type conversion and change the factor level of Listing
Category. Afterwards we look at the summary of all variables of interest. 
Almost all the values are in normal range, only in investors we see loans funded
by 1189 members to the maximum, which is an unusual thing.

```{r echo=FALSE, Univariate_Plots_1}

# change column type from integers to factors
df.loan$Term <- factor(df.loan$Term)
df.loan$ListingCategory <- factor(df.loan$ListingCategory..numeric.)
df.loan$ProsperScore = factor(df.loan$ProsperScore)
df.loan$LoanOriginationDate <- as.POSIXct(df.loan$LoanOriginationDate,
                                          format = "%Y-%m-%d %H:%M:%OS")

# re-order factor levels for IncomeRange
df.loan$IncomeRange <- factor(df.loan$IncomeRange, 
                           levels = c("$0", "$1-24,999", "$25,000-49,999",
                                      "$50,000-74,999", "$75,000-99,999",
                                      "$100,000+", "Not employed", "Not displayed"))

# change factor level values for ListingCategory
df.loan$ListingCategory <- revalue(df.loan$ListingCategory, 
                                   c("0"="Not Available", "1"="Debt Consolidation",
                                     "2"="Home Improvement", "3"="Business",
                                     "4"="Personal Loan", "5"="Student Use",
                                     "6"="Auto", "7"="Other",
                                     "8"="Baby&Adoption","9"="Boat",
                                     "10"="Cosmetic Procedure", "11"="Engagement Ring", 
                                     "12"="Green Loans", "13"="Household Expenses", 
                                     "14"="Large Purchases", "15"="Medical/Dental", 
                                     "16"="Motorcycle", "17"="RV", 
                                     "18"="Taxes", "19"="Vacation", 
                                     "20"="Wedding Loans"))


# select subset of the complete dataset for generating plots
selected.features <- c("Term", "LoanStatus", "BorrowerRate", "ListingCategory", 
                       "CreditScoreRangeLower", "CreditScoreRangeUpper",
                       "IncomeRange", "LoanOriginalAmount", "LoanOriginationDate",
                       "PercentFunded", "ProsperScore", "Recommendations", 
                       "Investors")
df.loan.subset <- df.loan[, selected.features]
summary(df.loan.subset)
```
# Univariate section
#### Is the number of loans affected by choices of different interest rate models?
As mentioned earlier, Prosper used two models for calculating loan interest rate.
The plot below shows a drop in year 2009, and this is exactly when the new 
interest rate model was applied. Afterwards the number of loans started increasing
steadily every year.
```{r echo=FALSE, Univariate_Plots_2}
df.loan.subset$LoanOriginationYear <- format(df.loan.subset$LoanOriginationDate,
                                             "%Y")
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginationYear)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  scale_y_continuous() +
  ggtitle("Number of loans per year from 2005 to 2014")

```

#### What is the most chosen term of loans?
The plot below shows most people have chosen loans of longer terms. 36 months of
loans are chosen by most of the people.
```{r echo=FALSE, Univariate_Plots_3}
ggplot(data = df.loan.subset, mapping = aes(x = Term)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  scale_y_continuous(breaks = seq(0, 100000, 10000)) +
  ggtitle("Number of loans per term")

```

#### What are the income range of borrowers?
One interesting question is which income level contributes to the most number of
loans. This helps us understand the target user base and provide more customized
services in the future. I thought people with low income would have most number 
of loans, because people would need more financial help to satisfy some basic 
living requirements such as buying a house or education for children. Contrarily, 
this plot shows people with income 25000 to 75000 contributed to the most number 
of loans, instead of people with income 1 to 24999. Also, it is surprising to see
people with income 0 could actually successfully have a loan. One guess might be
the income only contains income from a job, instead of other sources.
```{r echo=FALSE, Univariate_Plots_4}
# change the order level of the factor
ggplot(data = df.loan.subset, mapping = aes(x = IncomeRange)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Number of loans per income range")

```

#### For what purposes do people borrow loans?
The plot below shows different purposes of loans. The category of debt-consolidation
has most number of loans. Given in this case people borrow loan here to pay off
other loans, it indicates that the loan interest rate would be lower than from
other sources. We will look into detail when analyzing the purpose vs. loan
interest rate later.
```{r echo=FALSE, Univariate_Plots_5}
ggplot(data = df.loan.subset, mapping = aes(x = ListingCategory)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  scale_y_continuous(breaks = seq(0, 100000, 10000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Number of loans per category")

```

#### What is the range of loan interest rate?
Next we generate a histogram for borrower interest rate with bin width 0.01. 
Overall, the most frequent interest rate is around 0.15, with frequency decreasing
along each side. It is worth noticing in the range of 0.32-0.33, there is another 
peak. This might be some default value determined by the system. Later we can 
answer this question by looking at average interest rates over time.
```{r echo=FALSE, Univariate_Plots_6}
qplot(data = df.loan.subset, 
      x = BorrowerRate, 
      geom ="histogram", 
      binwidth = 0.01,
      main = "Histogram for borrower interest rate",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3))

```

#### How much money will people borrow in a loan?
This histogram shows the loan amount. From the histogram we can see a few peaks.
Many loans are centered around bins in 4000, 5000, 10000, 15000, 20000 and 25000.

```{r echo=FALSE, Univariate_Plots_7}
qplot(data = df.loan.subset, 
      x = LoanOriginalAmount, 
      geom ="histogram", 
      binwidth = 500,
      main = "Histogram for loan amount bin=500",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3)) +
  scale_x_continuous(breaks=seq(0, 40000, by=5000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

If we change the bin size to 5000, then the discrete peaks would disappear. 
Now the plot shows a more smooth result with most of the loans falling into
5000-10000 category.

```{r echo=FALSE, Univariate_Plots_8}
qplot(data = df.loan.subset, 
      x = LoanOriginalAmount, 
      geom ="histogram", 
      binwidth = 5000,
      main = "Histogram for loan amount bin=5000",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3)) +
  scale_x_continuous(breaks=seq(0, 40000, by=5000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### How are the prosper scores distributed across borrowers?
The plot below shows that there are many missing values. Apart from that borrowers'
score are almost normally distributed. Most of the borrowers have score ranging 
from 4 to 8. 

```{r echo=FALSE, Univariate_Plots_9}
ggplot(data = df.loan.subset, mapping = aes(x = ProsperScore)) + 
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  ggtitle("Borrowers' prosper scores")
```

#### How are the credit ratings distributed across borrowers?
The plot below shows that the lower and upper bound of credit ratings are approximately
within the same range.

```{r echo=FALSE, Univariate_Plots_10}
p1 <- ggplot(data = df.loan.subset, mapping = aes(x = CreditScoreRangeLower)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  ggtitle("Credit scores lower bound")
p2 <- ggplot(data = df.loan.subset, mapping = aes(x = CreditScoreRangeUpper)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  ggtitle("Credit scores upper bound")
grid.arrange(p1, p2, nrow = 2)
```

#### How many people will invest per loan?
This histogram shows the number of investors for the loans. Most of the loans are 
invested by 1-10 investors. 

```{r echo=FALSE, Univariate_Plots_11}
qplot(data = df.loan.subset, 
      x = Investors, 
      geom ="histogram", 
      binwidth = 10,
      main = "Histogram for number of investors",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3))

```

# Bivariate Plots Section

#### For all the selected numerical variables, how are they correlated?
From the correlation table we can see the loan original amount and investors have
a slight positive correlation. This is not surprising, since larger amount of 
loans would be easier to be invested with multiple people rather than only one
person. Also the loan original amount and borrower rate are slightly negative 
correlated, i.e. the more money people borrow, the less interest rate, this is
also not an unusual thing.

```{r echo=FALSE, Bivariate_Plots_1}
nums <- sapply(df.loan.subset, is.numeric)
cor(df.loan.subset[, nums])

```

#### How are the loan amount distributed for different income categories?
The plot below shows people with higher income borrows a larger amount of loan.
Again the unexpected thing happened with 0 income people. Not only they were able
to borrow a loan, some of them can actually borrow a large amount of loan, as we
can see from the outliers quite a few loan amount are between 22000 to 25000.

```{r echo=FALSE, Bivariate_Plots_2}
ggplot(data = df.loan.subset, mapping = aes(x = IncomeRange,
                                            y = LoanOriginalAmount,
                                            group = IncomeRange)) + 
  geom_boxplot(aes(fill=IncomeRange)) +
  scale_fill_brewer(palette="Blues") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Loan amount for different income levels")

```

#### How are the loan interest distributed for different income categories?
Overall we see a decreasing trend with people's income getting higher. One 
particular category to notice is the unemployed category rate. For unemployed
people, the rates are slightly higher than other income categories (~0.25). Given
previously we also noticed higher income people tend to borrow more money, and 
the larger amount of loan results in lower interest rate, it makes sense to see
this trend here.

```{r echo=FALSE, Bivariate_Plots_3}
ggplot(data = df.loan.subset, mapping = aes(x = IncomeRange,
                                            y = BorrowerRate,
                                            group = IncomeRange)) + 
  geom_boxplot(aes(fill=IncomeRange)) +
  scale_fill_brewer(palette="Blues") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Loan interest rate for different income levels")

```

#### How are the loan interest distributed for different loan purposes?
Overall we don't see any category to have a particularly high average loan rate. 
The average rates range from 0.15 to 0.25.

```{r echo=FALSE, Bivariate_Plots_4}
ggplot(data = df.loan.subset, mapping = aes(x = ListingCategory,
                                            y = BorrowerRate,
                                            group = ListingCategory)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Loan interest rate for different listing categories")

```

#### How are the loan interest distributed for different prosper ratings?
Overall we see the trend of higher rating with a lower average of loan interest
rate. This helps us understand what is important in determining the interest rate.

```{r echo=FALSE, Bivariate_Plots_5}
ggplot(data = df.loan.subset, mapping = aes(x = ProsperScore,
                                            y = BorrowerRate,
                                            group = ProsperScore)) + 
  geom_boxplot() +
  ggtitle("Loan interest rate for different prosper ratings")

```

#### How are the loan interest rate distributed for number of investors?
Here we could see a slightly decreasing trend for the number of investors when 
the borrower rate goes high. 

```{r echo=FALSE, Bivariate_Plots_6}
ggplot(data = df.loan.subset, mapping = aes(x = BorrowerRate,
                                            y = Investors)) + 
  geom_point() +
  ggtitle("Loan interest rate vs number of investors")

```


# Multivariate Plots Section

#### How does the borrower interest rate for different terms change over time?
From the plot below we can see loan terms of 12 months and 60 months did not 
exist until somewhere end of 2010. The interest rate for 12 months loans are 
varying quite a lot, whereas the interest rate for 36 and 60 months loans are 
changing in a more steady manner.

```{r echo=FALSE, Multivariate_Plots_1}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginationDate,
                                            y = BorrowerRate,
                                            color = Term,
                                            alpha = 0.1)) + 
  stat_summary(fun.y="mean", geom = "point") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Average loan interest rate for different loan terms")
```

#### How does the borrower interest rate for people with different prosper 
scores change overtime?
From the plot we can see there is no user with score 11 before 2013. As for the 
rest, the interest rate is going down as time goes. Again, people with higher 
prosper scores have a lower rate for loans than people with low scores.

```{r echo=FALSE, Multivariate_Plots_2}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginationDate,
                                            y = BorrowerRate,
                                            color = ProsperScore)) + 
  facet_wrap( ~ ProsperScore) +
  stat_summary(fun.y="mean", geom = "point", alpha = 0.3) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Average loan interest rate for different prosper scores overtime")
```


------

# Final Plots and Summary

In this project I mainly explored the relationships between borrowers' interest
rate and borrowers' income, prosper rating and term respectively. 

Overall we see, with people's income increasing, the average interest rate is 
decreasing. In particular we noticed the rate for unemployment category. The rates 
are slightly higher than other income categories. 

```{r echo=FALSE, Plot_One}
ggplot(data = df.loan.subset, mapping = aes(x = IncomeRange,
                                            y = BorrowerRate,
                                            group = IncomeRange)) + 
  geom_boxplot(aes(fill=IncomeRange)) +
  scale_fill_brewer(palette="Blues") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Loan interest rate for different income levels")

```

From the plot below we can see after applying the new rate calculation model,
Prosper also provided additional service types, i.e. 12 month and 60 month loan.
Also, the interest rate for 12 months loans are varying in a more unpredictable
manner compared with the rates for 36 and 60 months loans.


```{r echo=FALSE, Plot_Two}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginationDate,
                                            y = BorrowerRate,
                                            color = Term)) + 
  stat_summary(fun.y="mean", geom = "point", aes(alpha = 0.3)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Average loan interest rate for different loan terms")
```

Lastly, we see the importance of prosper scores in correlation with the interest
rate. It is obvious that as the prosper scores grow higher, the average interest
rate is lower. Also, as time goes the interest rate shows a decreasing trend.

```{r echo=FALSE, Plot_Three}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginationDate,
                                            y = BorrowerRate,
                                            color = ProsperScore)) + 
  facet_wrap( ~ ProsperScore) +
  stat_summary(fun.y="mean", geom = "point", alpha = 0.3) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Average loan interest rate for different prosper scores overtime")
```

------

# Reflection

The most struggling part is to select the relevant variables in the dataset.
There are too many variables and without a focus or a list of questions in mind,
it is really difficult to make a decision. In addition, understanding the dataset
is also difficult without a proper context. After getting a good understanding 
of how Prosper is running its business and how the peer-to-peer lending works, 
the analysis got easier to proceed. The biggest lesson learnt here is, there could
not be a good analysis without a good understanding of the business / industry.
