Loan analysis by Xin Pang
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

packages <- c("dplyr", "ggplot2")
new.packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new.packages)){
  install.packages(new.packages)
}
library(dplyr)
library(ggplot2)
library(plyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df.loan <- read.csv("./prosperLoanData.csv", header = TRUE)
```

**Background**
Prosper.com is a peer-to-peer lending website privoding individuals with 
services to borrow money or invest in personal loans. According to [Wikipedia](https://en.wikipedia.org/wiki/Prosper_Marketplace) this 
company used two different ways to calculate loan interest rates. From 2006 to 
2009 the loan interest rate was determined by borrowers and lenders in a bidding
system. Afterwards the rate is predefined by Prosper based on a formula 
evaluating each prospective borrower's credit risk. The dataset to be analyzed
contains users' loan details from November 2005 to March 2014, including 113937 
loan records with 81 variables.

## Overview

We want to know the type of each column, wether they are numeric or factor, 
such that we can select a few of them for analysis. Due to the large number of 
variables in the dataframe, here I only show the type of first 10 attributes.
From the result we can see there are both factor variables and numerical variables.

```{r echo=FALSE, Univariate_Plots_0}
head(lapply(df.loan, class), n = 10)
```

The main features of interest are the following:

* Term: The length of the loan expressed in months. (integer->factor)
* Loan status: The current status of the loan. (factor)
* Borrower rate: The borrower's interest rate for this loan. (numeric)
* Listing category: The category of the listing that the borrower selected when 
posting their listing. (integer->factor)
* Income range: The income range of the borrower at the time the listing was 
created. (factor)
* Loan original amount: The origination amount of the loan. (integer)
* Loan origination date: The origination date of the loan. (factor->date)
* Percent funded: Percent the listing was funded. (numeric)
* Recommendations: Number of recommendations the borrower had at the time the 
listing was created. (integer)
* Investors: The number of investors that funded the loan. (integer)

There are a few variables requiring a conversion for proper types.

* Term: contains 3 values and can be changed to a factor.
* Listing category: they are actually coded into numbers, each number 
representing one category, thus it should be factor instead of integer.
* Loan origination date: should be in date format.

Below processes the column type conversion and change the factor level of Listing
Category. Afterwards we look at the summary of all variables of interest. 
Almost all the values are in normal range, only in investors we see loans funded
by 1189 members to the maximum, which is an unusual thing.

```{r echo=FALSE, Univariate_Plots_1}

# change column type from integers to factors
df.loan$Term <- factor(df.loan$Term)
df.loan$ListingCategory <- factor(df.loan$ListingCategory..numeric.)
df.loan$LoanOriginationDate <- as.POSIXct(df.loan$LoanOriginationDate,
                                          format = "%Y-%m-%d %H:%M:%OS")

# re-order factor levels for IncomeRange
df.loan$IncomeRange <- factor(df.loan$IncomeRange, 
                           levels = c("$0", "$1-24,999", "$25,000-49,999",
                                      "$50,000-74,999", "$75,000-99,999",
                                      "$100,000+", "Not employed", "Not displayed"))

# change factor level values for ListingCategory
df.loan$ListingCategory <- revalue(df.loan$ListingCategory, 
                                   c("0"="Not Available", "1"="Debt Consolidation",
                                     "2"="Home Improvement", "3"="Business",
                                     "4"="Personal Loan", "5"="Student Use",
                                     "6"="Auto", "7"="Other",
                                     "8"="Baby&Adoption","9"="Boat",
                                     "10"="Cosmetic Procedure", "11"="Engagement Ring", 
                                     "12"="Green Loans", "13"="Household Expenses", 
                                     "14"="Large Purchases", "15"="Medical/Dental", 
                                     "16"="Motorcycle", "17"="RV", 
                                     "18"="Taxes", "19"="Vacation", 
                                     "20"="Wedding Loans"))


# select subset of the complete dataset for generating plots
selected.features <- c("Term", "LoanStatus", "BorrowerRate", "ListingCategory", 
                       "IncomeRange", "LoanOriginalAmount", "LoanOriginationDate",
                       "PercentFunded", "Recommendations", "Investors")
df.loan.subset <- df.loan[, selected.features]
summary(df.loan.subset)
```

#### Is the number of loans affected by choices of different interest rate models?
As mentioned earlier, Prosper used two models for calculating loan interest rate.
The plot below shows a drop in year 2009, and this is exactly when the new 
interest rate model was applied. Afterwards the number of loans started increasing
steadily every year.
```{r echo=FALSE, Univariate_Plots_2}
df.loan.subset$LoanOriginationYear <- format(df.loan.subset$LoanOriginationDate,
                                             "%Y")
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginationYear)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  scale_y_continuous() +
  ggtitle("Number of loans per year from 2005 to 2014")

```

#### What is the most chosen term of loans?
The plot below shows most people have chosen loans of longer terms. 36 months of
loans are chosen by most of the people.
```{r echo=FALSE, Univariate_Plots_3}
ggplot(data = df.loan.subset, mapping = aes(x = Term)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  scale_y_continuous(breaks = seq(0, 100000, 10000)) +
  ggtitle("Number of loans per term")

```

#### What are the income range of borrowers?
One interesting question is which income level contributes to the most number of
loans. This helps us understand the target user base and provide more customized
services in the future. I thought people with low income would have most number 
of loans, because people would need more financial help to satisfy some basic 
living requirements such as buying a house or education for children. Contrararily, 
this plot shows people with income 25000 to 75000 contributed to the most number 
of loans, instead of people with income 1 to 24999.
```{r echo=FALSE, Univariate_Plots_4}
# change the order level of the factor
ggplot(data = df.loan.subset, mapping = aes(x = IncomeRange)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Number of loans per income range")

```

#### For what purposes do people borrow loans?
The plot below shows different purposes of loans. The category of debt-consolidation
has most number of loans. Given in this case people borrow loan here to pay off
other loans, it indicates that the loan interest rate would be lower than from
other sources. We will look into detail when analyzing the purpose vs. loan
interest rate later.
```{r echo=FALSE, Univariate_Plots_5}
ggplot(data = df.loan.subset, mapping = aes(x = ListingCategory)) +
  geom_bar(color = "white", fill = "blue", alpha = 0.3) +
  scale_y_continuous(breaks = seq(0, 100000, 10000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Number of loans per category")

```
#### What is the range of loan interest rate?
Next we generate a histogram for borrower interest rate with bin width 0.01. 
Overall, the most frequent interest rate is around 0.15, with frequency decreasing
along each side. It is worth noticing in the range of 0.32-0.33, there is another 
peak. This might be some default value determined by the system. Later we can 
answer this question by looking at average interest rates over time.
```{r echo=FALSE, Univariate_Plots_6}
qplot(data = df.loan.subset, 
      x = BorrowerRate, 
      geom ="histogram", 
      binwidth = 0.01,
      main = "Histogram for borrower interest rate",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3))

```

#### How many people will invest per loan?
This histogram shows the number of investors for the loans. Most of the loans are 
invested by 1-10 investors. 
```{r echo=FALSE, Univariate_Plots_7}
qplot(data = df.loan.subset, 
      x = Investors, 
      geom ="histogram", 
      binwidth = 10,
      main = "Histogram for number of investors",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3))

```
#### How much money will people borrow in a loan?
This histogram shows the loan amount. From the histogram we can see a few peaks.
Many loans are centered around bins in 4000, 5000, 10000, 15000, 20000 and 25000.
```{r echo=FALSE, Univariate_Plots_8}
qplot(data = df.loan.subset, 
      x = LoanOriginalAmount, 
      geom ="histogram", 
      binwidth = 500,
      main = "Histogram for loan amount bin=500",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3)) +
  scale_x_continuous(breaks=seq(0, 40000, by=5000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
If we change the bin size to 5000, then the discrete peaks would disappear. 
Now the plot shows a more smooth result with most of the loans falling into
5000-10000 category.
```{r echo=FALSE, Univariate_Plots_8}
qplot(data = df.loan.subset, 
      x = LoanOriginalAmount, 
      geom ="histogram", 
      binwidth = 5000,
      main = "Histogram for loan amount bin=5000",
      fill = I("blue"),
      col = I("white"),
      alpha = I(.3)) +
  scale_x_continuous(breaks=seq(0, 40000, by=5000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


# Bivariate Plots Section

> In the bi-variate plot section, we are mainly interested in the following 
questions.

* How does the loan amount affect the number of investors? If the loan amount
is large, will more investors invest?
* How does the loan amount affect the percentage of money funded? If the loan 
amount is large, will it be less likely to have a completely funded loan?
* Will people from lower income range borrow more money?

```{r echo=FALSE, Bivariate_Plots_1}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginalAmount, 
                                            y = Investors)) +
  geom_point()

```
```{r echo=FALSE, Bivariate_Plots_2}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginalAmount, 
                                            y = PercentFunded)) +
  geom_point() +
  ggtitle("Percent funded ~ loan amount")

```

```{r echo=FALSE, Bivariate_Plots_3}
ggplot(data = df.loan.subset, mapping = aes(x = LoanOriginalAmount, 
                                            y = Investors)) +
  geom_point()

```

```{r}
cor(df.loan.subset$LoanOriginalAmount, df.loan.subset$Investors)
cor(df.loan.subset$LoanOriginalAmount, df.loan.subset$BorrowerRate)

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!
